\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2023-03-10 sections/chameleon.tex
Latest
Squeeze

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error in Fig.~\ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but it can be more problematic in large programs where the lines contributing to the type error are far apart in the source code.

@@

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of a complex computation. But the details of this process are hard to explain to users and are usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations. These have to be rediscovered by programmers using less rigorous methods.

@@

\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'.

@@

\chameleon{} comprises two parts: a type inference engine and a novel interactive debugging interface. This separation of concerns allows us to (1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications; and (2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine. The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements, as described in Section \ref{sec:typeinferenceengine}.

@@

\begin{table}
    \centering
\begin{scriptsize}
\begin{center}
    \begin{tabular}{ l l  }
     \textit{Mode} & \textit{Features} \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Card \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Card \\
     & Deduction Steps \\
    \end{tabular}
    \end{center}
\end{scriptsize}
    \caption{\chameleon{} modes and features}
    \label {tab:chameleon-features}
\end{table}

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\begin{listing}[!ht]
\begin{minted}{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace{-3mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in Listing~\ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS. An example output for Listing~\ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
    \caption{
    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
    }
    \label{fig:motivation-example}
\end{figure}
\subsubsection{\textbf{Traditional type errors show only limited location}}
Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error in Fig.~\ref{fig:motivation-example} can be fixed in multiple locations. For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}. In the type error message, only the \texttt{addPair} expression on line 4 was blamed.  In this small example, the whole context is visible, but it can become problematic in large programs where the lines contributing to the type error are far apart in the source code.

@@

\subsubsection{\textbf{Traditional type errors give poor explanations}}
When the compiler rejects a program, the internal state of type checking is the result of a complex computation. But the details of this process are hard to explain to users and are usually not reported by compilers. For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous declarations. These have to be rediscovered by programmers using less rigorous methods.

@@

\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is `actual' or `expected'.

@@

\chameleon{} comprises two parts: a type inference engine and a novel interactive debugging interface. 
% This separation of concerns allows us to (1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications; and (2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine. 
The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements, as described in Section \ref{sec:typeinferenceengine}.

@@

\begin{table}
    \centering
\begin{scriptsize}
\begin{center}

    \begin{tabular}{ l l  }
     \textit{Mode} & \textit{Features} \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Cards \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Cards \\
     & Deduction Steps \\
    \end{tabular}
    \end{center}
\end{scriptsize}
    \caption{\chameleon{} modes and features}
    \label {tab:chameleon-features}
\end{table}

@@

% \begin{lstlisting} [
%     language=Haskell, caption={
%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
% }, label={listing:ifelse}
% ]
% if a then b else c
% a = "True"
% \end{lstlisting}
\definecolor{bg}{rgb}{0.95,0.95,0.95}
\begin{listing}[!ht]
\begin{minted}[bgcolor=bg,linenos]{haskell}
    if a then b else c
    a = "True"
\end{minted}
\vspace*{-9mm}
\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
\label{listing:1}
\end{listing}
For instance, for the program in Listing~\ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS. An example output for Listing~\ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}. This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}). \looseness=-1

### 2023-03-10 sections/discussion.tex
Squeeze

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examines   the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using the \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use \chameleon{}'s interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than passively reading the type error output. In this section, we will discuss a few interpretations of the results.

\subsection{Effect on Reading Source Code}
From the results of Study 1a, we observed that the choice of debugging tool had little effect on how fast programmers solve simple type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered) in study 1b, programmers are more effective using \chameleon{}. One explanation is that \chameleon{} reduces the amount of reading time by taking programmers more directly to the problem. Earlier studies \cite{jbara_how_2015, peitek_what_2020} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, these may be incomplete, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

@@

\begin{listing}
\begin{minted}{haskell}
f z     
    | z == 3 = False
    | z == '4' = True
\end{minted}
\caption{In this simple program, \chameleon{} reports the error happened in \texttt{f} and \texttt{z}.}
\label{listing:2}
\end{listing}
% Implication for research
% In our studies, \chameleon{} shows a few different designs for type error visualization and interaction. However, we are far from exhausting the search space. One challenge we notice in the current design is that programmers have to shift their focus between the editor on the left and the \chameleon{} debugging interface on the right. Using the hover popup window in most mainstream IDEs may reduce the context switching during debugging.

>>>>

This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examines   the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use \chameleon{}'s interactive features are more efficient in fixing type errors than passively reading the type error output. In this section, we will discuss a few interpretations of the results.

\subsection{Effect on Reading Source Code}
From the results of Study 1a, we observed that the choice of debugging tool had little effect on how fast programmers solve simple type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered) in study 1b, programmers are more effective using \chameleon{}. One explanation is that \chameleon{} reduces the amount of reading time by taking programmers more directly to the problem. Earlier studies \cite{jbara_how_2015, peitek_what_2020} showed that reading source code is generally the initial step of solving programming problems and is done in several passes. Although traditional compiler error message tools initially show fewer locations, these may be incomplete, meaning that programmers have to expand the reading span without clear guidance. In contrast, \chameleon{} shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

@@

\begin{listing}
\begin{minted}[bgcolor=bg,linenos]{haskell}
f z     
    | z == 3 = False
    | z == '4' = True
\end{minted}
\vspace*{-9mm}
\caption{A simple Haskell program with a type error}
\label{listing:2}
\end{listing}
% Implication for research
% In our studies, \chameleon{} shows a few different designs for type error visualization and interaction. However, we are far from exhausting the search space. One challenge we notice in the current design is that programmers have to shift their focus between the editor on the left and the \chameleon{} debugging interface on the right. Using the hover popup window in most mainstream IDEs may reduce the context switching during debugging.

### 2023-03-10 sections/evaluation.tex
Squeeze

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users, and the individual differences among participants are generally more significant than the differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group. We attribute this significant difference to the difficulty of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we might observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

@@

\textbf{RQ2.1}. The most striking feature of the data is that users tend to vary wildly in their use of the tool. Some users used the features extensively, while others completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups in table~\ref{tab:interaction-level}.
\begin{table}
    \centering
\begin{scriptsize}
\begin{small}
\noindent\begin{tabularx}{\linewidth}{ 
  | >{\hsize=.3\hsize}X 
  | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }
    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the low interaction group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}
\end{scriptsize}
    \caption{Levels of programmer interaction and their description}
    \label {tab:interaction-level}
\end{table}

@@

Another observation is when using the mode switching feature of \chameleon{}, we show this by presenting the starting mode and finishing mode of each task and each participant in a correlation matrix (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes: 53\% changing modes vs. 47\% staying in the same mode. Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

>>>>

The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7. In hindsight, these tasks were trivial challenges for most users, and the individual differences among participants are generally more significant than the differences between treatments. However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group. We attribute this significant difference to the difficulty of Task 8. The source file is longer and involves more language features (abstract data types and high-level functions). GHC struggles to produce a relevant error message for this type of error. 
From this result, we hypothesized that we might observe a more significant difference using tasks with lengthier and more realistic source code. 
This hypothesis is also supported by the most common feedback that the tasks were too trivial to invite meaningful evaluation. One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

@@

\textbf{RQ2.1}. The most striking feature of the data is that users tend to vary wildly in their use of the tool. Some users used the features extensively, while others completed the tasks without actively exploring the given information. Based on this discrepancy, we divided the users into three groups in table~\ref{tab:interaction-level}.
\begin{table}
    \centering
\begin{scriptsize}
\begin{small}
\noindent\begin{tabularx}{\linewidth}{ 
  | >{\hsize=.26\hsize}X 
  | >{\hsize=.74\hsize \raggedright\arraybackslash}X  | }
    \hline
        Interaction level & Description \\ \hline
        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        \textit{High}  & Users did everything from the low interaction group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards. \\ \hline
\end{tabularx}
\end{small}
\end{scriptsize}
    \caption{Levels of programmer interaction and their description}
    \label {tab:interaction-level}
\end{table}

@@

Another observation is when using the mode switching feature of \chameleon{}, we show this by presenting the starting mode and finishing mode of each task and each participant in a correlation matrix (Fig. \ref{fig:r4-mode-switching}). This observation suggests two characteristics of using multi-mode debugging tools. First, to answer \textbf{RQ2.2} programmers are roughly splitted in this matter: 53\% changing modes vs. 47\% staying in the same mode. Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of the more concise ones.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
        % The rows represent the finishing mode, columns starting modes.
    }
    \label{fig:r4-mode-switching}
\end{figure}

### 2023-03-10 sections/related-work.tex
Squeeze

\subsection{Finding all type error locations}
Many have studied the approach of finding all locations that contribute to a type error~\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}. Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used  Constraint handling rules (CHR) to support the computing of type error slices in Haskell. Chameleon also supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.
To the best of our knowledge, ours is the first user-centered evaluation of an interactive type debugging system involving type-error slicing.

>>>>

\subsection{Finding all type error locations}
Many have studied the approach of finding all locations that contribute to a type error~\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}. Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and Minimal Unsatisfiable Subset (MUS) manipulation are used to generate these slices. The language supported in Haack's work was a subset of Standard ML. The original Chameleon~\cite{stuckey_interactive_2003} used  Constraint handling rules (CHR) to support the computing of type error slices in Haskell. Chameleon also supported advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line interface. Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies. While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.
To the best of our knowledge, ours is the first user-centered evaluation of an interactive type debugging system involving type-error slicing.
